Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2019-12-17 write-up/fda-recurrent.tex
Update fda-recurrent.tex

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. We use the sandwich smoother~\citep{Xiao2013} to estimate the mean~$\mu_y(s,t) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009} bivariate tensor product splines~\citep{Wood2006} and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(s,t) = X(s,t) - \hat \mu_y(s,t)$ to be the mean-zero process at each time~$t \in \bfT_i \cup D_i$.

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma_y (s, s^\prime) = \int_{0}^\tau c_y( (s,T), (s^\prime, T) ) g_y(T) dt.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c_y((s,t), (s^\prime,t))$ is the covariance function of the windowed covariate history~$X(t,\cdot)$ and $g_y(T)$ is the intensity function for event ($y=1$) and subsampled ($y=0$) times respectively. Estimation of~$\Sigma_y$ occurs in two steps. For simplicity, we present the steps for subsampled times (i.e., $y=0$) but the steps are the same for event times as well. First, the pooled sample covariance is calculated at a set of grid points:
\[
\tilde \Sigma_0 (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | \bfD_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in \bfD_i} \tilde X (s_r,t) \tilde X (s_{r^\prime}, t) \right).
\]
Due to our concern over independent evolution as discussed in section~\ref{section:memproblems}, we do not assume that each observation is observed with white noise and therefore the diagonal
elements of~$\hat \Sigma_0$ are not inflated. Second, the estimator~$\hat \Sigma$ is further smoothed again using the sandwich smoother~\citep{Xiao2013}. Note~\cite{Park2018} smooth the off-diagonal elements, while here we smooth the entire pooled sample covariance matrix. All negative eigenvalues are set to zero to ensure positive semi-definiteness. The result is used as an estimator~$\hat \Sigma_0$ for the pooled covariance $\Sigma_0$.

@@

The connection to logistic regression allows our methodology to leverage ``off-the-shelf'' software.  The main complication is pre-processing of the functional data; however, these additional steps can also be taken care of via existing software.  Therefore, the entire data analytic pipeline is easy-to-implement and require minimal additional effort by the end-user. Before demonstrating the methodology via simulation in Section~\ref{section:simstudy}) and a worked example in Section~\ref{section:example}, we detail a theoretical analysis of our proposal.

@@

Our theoretical analysis requires assumptions regarding the subsampling procedure, the event process, and the functional data. We state these assumptions and then our main theorems. We start by assuming there exists a~$\tau < \infty$ such that all individuals are no longer at risk (i.e.,~$\tau_i < \tau$ for all~$i$). Moreover, define $R_i (t)$ to be the at risk indicator for participant~$i$, i.e., $R_i (t) = {\bf 1} [t \in (0,\tau_i)]$. Asymptotic theory provided in Lemma~\ref{lemma:simpleasym} will be proven under regularity conditions A-E in~\cite[pp. 420--421]{Andersen1993} along with the following additional assumptions:

\begin{assumption}[Event process assumptions]
\label{assumption:events}\normalfont
We assume the following holds:
\begin{enumerate}[label=(E.\arabic*)]
\item\label{E1} The subsampling rate is both lower and upper
  bounded for all ``at risk times''; that is,~$0 < L < \pi_i (t) < U <
  \infty$ for all~$i=1,2,\ldots$ and~$t \in [0,\tau]$.
\item\label{E2} There exists a nonnegative definite matrix~$\Xi
  (\theta)$ such that
  \[
    n^{-1} \Xi_n (\theta) = n^{-1} \sum_{i=1}^n \int_0^\tau w_i (t;
    \theta) (1 - w_i (t; \theta)) \left[ \frac{h_i^{(1)}(t;
        \theta)}{h_i (t; \theta)} \right] \times  \left [
      \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top R_i (t)
    dt \overset{P}{\to} \Xi (\theta).
  \]
\item\label{E3} There exists~$M$ such that $|W_{i,j,t}| < M$ for
  all~$(i,j,t)$.
\item\label{E4} For all~$j,k$
\[
n^{-1} \sum_{i=1}^n \int_0^\tau \left | \frac{d^2}{d\theta_j
    d\theta_k} h_i (t;\theta_0) \right|^2 R_i (t) dt \overset{P}{\to}
C < \infty
\]
as $n \to \infty$.
\end{enumerate}
\end{assumption}

We also require several assumptions due to the truncation of the Karhunen-Lo{\`e}ve decomposition that represents~$X(t,s)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(s, T) : (s,T) \in \mathcal{S} \times
  \mathcal{T} \}$ is a square integrable element of the $L^2 (
  \mathcal{S} \times \mathcal{T})$
\item\label{A2} The subsampling and conditional intensity rate functions~$g_y(T)$ are continuous
  and~$\sup |g_y(T)| < \infty$.
\item\label{A3} $\E[X(s,T) X(s^\prime, T) X(s,T^\prime)
  X(s^\prime, T^\prime) ] < \infty$ for
  each $s,s^\prime \in [0,\Delta]$ and~$0 < T, T^\prime < \tau$.
\item\label{A4} $\E[\|X(\cdot, T)\|^4] < \infty$ for each~$0< T
  < \tau$.
\end{enumerate}
\end{assumption}
Assumption~\ref{A3} is guaranteed by~\ref{E1} and~\ref{E3}, but is stated here to emphasize its importance for the marginal covariance procedure of~\cite{Park2018}. Finally, for simplicity, we assume that there exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the
true function~$\beta(t)$ sits in the span of the spline basis expansion. With the necessary assumptions stated, we can now state Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.
\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events},
Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for
large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
An estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i}
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator
for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the
previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process,
standard logistic regression software can be used to fit the recurrent
event model by specifying an offset equal to~$\log \pi_i (t)$.
Not only this, Lemma~\ref{prop:optimal} shows
weights~\eqref{eq:waage_weights} are optimal within a particular class
of weighted estimating equations.

\begin{prop} \normalfont
\label{prop:optimal}
If the event process is an inhomogeneous Poisson point process with
intensity~$h(t; \theta)$ and subsampling occurs via an independent,
inhomogeneous Poisson point process with intensity~$\pi (t)$,
then~$\hat U_n (\theta)$ are optimal estimating functions (i.e., most
efficient) in the class of weighted estimating functions given
by~\eqref{eq:approxscore} replacing~\eqref{eq:waage_weights} by any
weight function~$w_i (t; \theta)$. This class includes the
Horvitz-Thompson estimator under~$w(s; \theta) = 1$.
\end{prop}

\noindent Proposition~\ref{prop:optimal} ensures the only loss of
statistical efficiency is due to subsampling and not using a
suboptimal estimation procedure given subsampling.

@@

Under assumption~\ref{assumption:truncation}, we next consider the
statistical efficiency of our proposed estimator when compared to
complete-data maximum likelihood estimation.
While subsampling introduces additional variation, it may
significantly reduce the overall computational burden. It is this
trade-off that we next make precise.
In particular, we consider the following choice of subsampling
rate,~$\pi(t) = c \times h(t; \theta)$ for $c>0$. That is, the
subsampling rate is proportional to the intensity function with
time-independent constant~$c > 0$.
Under this subsampling rate, the weight
function~\eqref{eq:waage_weights} is equal to $c / (c+1)$.
Under Lemma~\ref{lemma:simpleasym},
\[
\Xi (\theta) = \frac{c}{c+1} \int_0^\tau \frac{ h^{(1)} (t; \theta)
  h^{(1)} (t; \theta)^\top}{h (t; \theta)} dt = \frac{c}{c+1} \Sigma (\theta)
\]
where~$\Sigma(\theta)$ is the Fisher information of the complete-data
maximum likelihood estimator.
Therefore the relative efficiency is~$c/(1+c)$. For an upper bound~$H
= \max_{t \in (0,\tau)} h(t;\theta)$, if we set~$\pi (t) = c \times
H$, then the relative efficiency can be lower bounded by~$c / (c+1)$.

Recall sensor measurements occur on the order of multiple times per
second.  Suppose the intensity rate is bounded above by $3$ and the
unit time scale was hours. If we can subsample the data at~$30$
times per hour, then we have a lower bound on the efficiency of
$0.909$. For a 4Hz sensor, this reduces the number of samples per hour
from~$4 \times 60 \times 60 = 14,400$ per hour to on average $30$.
While the computational complexity of logistic regression is linear in
the number of samples, we get $480$ times reduction in the data size
at the cost of a $0.909$ statistical efficiency. If we sample~$300$
times per hour, then the efficiency loss is only $0.999$, with a $48$
times reduction in data size. Table~\ref{tab:compvseff} provides
additional examples for a 4Hz sensor.  The data reduction depends on
the sensing rate; however, the lower bound on statistical efficiency
does not.  This is because the subsampling rate only depends on the
upper bound of the intensity function. That is, if the events are rare
then the amount of data we need to subsample is greatly reduced with
no impact to statistical efficiency.

@@

Recall theoretical results were proven under the assumption that there
exists~$b^\star$ such that~$\beta(t) = \phi(t) {\bf b}^\star$.
To make this assumption plausible, we set~$K_b$ large enough (but less
than~$K_x$ to ensure identifiability) such that the spline basis
expansion is sufficiently expressive. However, in practice, such a
choice of~$K_b$ may lead to overfitting the data. We therefore, in the
remainder of this paper, follow~\cite{GoldSmith2015} and induce
smoothing by assuming that~$\{ b_k \}_{k=3}^{K_b} \sim N(0, \sigma_b^2
I)$. Incorporating penalization is simple due to the connection
between with the score functions for logistic regression.

@@

Due to the connection with generalized mixed effects models, we can
leverage existing inferential machinery to obtain variance-covariance
estimates of model parameters.
That is, if~$\hat \Sigma_{bb}$ is the $K_b \times K_b$ deimnsional
matrix obtained by plugging in the REML estimates of variance
components into the formula for the variance of~$\hat b$, then the
standard error for estimate at time~$t_0$ -- i.e., ~$\hat \beta (t_0)
= \phi(t_0) \hat {\bf b}$ -- is given by~$\sqrt{ \phi (t_0 )
  \hat \Sigma_{bb} \phi(t_0)^\top}$.  Then the approximate 95\%
confidence interval can be constructed as~$\hat \beta (t_0) \pm 1.96
\sqrt{\phi (t_0) \hat \Sigma_{bb} \phi(t_0)^\top}$.

@@

We next assess the proposed methodology via a simulation study.
Simulation studies of joint high frequency sensor and recurrent
event data can be quite difficult to construct.  Here, we use the
imputation strategy introduced in Section~\ref{section:missingdata} to
circumvent some of these issues.

Here, we assume each individual is observed over five 12-hour days.
We then generate a sequence of event times via a marginal hazard
rate that ignores sensor data.  We assume the marginal rate is
constant and set to 1 per 12/6/1 hour(s). We then sample non-events at
rates of 1 per 12/6/1/0.5 hours.
This sampling strategy yields (non-)event times~$t$ along with
labels~$Y_t \in \{0,1\}$.
We then use the imputation strategy given by
equation~\eqref{eq:impute} to generate the curves~$\{ X(s,t) \}$.
For simplicity, we do not deal with
We assume that~$Y_t = 0$ implies the curve is flat with mean 0.
We assume that~$Y_t = 1$ implies three curves (A) the curve jumps
around 10 minutes prior to the event and remains elevated, (B) the
curve goes down then up and down smoothly, (C)
These curves are given in Figure~\ref{} for three choices of
parameters that indicate varying degrees of departure from the~$Y_t =
0$ case.
The associated~$\beta$ depends on the covariance~$\Sigma_t (s,t)$.
In particular, $\beta = \Sigma_t^{-1} (\mu_{1,t} - \mu_{0,t})$ is
guaranteed by Lemma~\ref{} and Theorem 1 of \cite{Petrovich2018}.
For~$\beta(s)$ non time-varying implies the covariance term

Using~\eqref{eq:}
We define~$X(t,s)$ at the grid of observerations per secondw ith
\[
  \Sigma_t (s, s^\prime) =
  \frac{\sigma^2}{\Gamma (v) 2^{\nu-1}}
  \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)^{\nu}
  K_v \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)
\]
where~$K_v$ is the modified Bessel function of the second kind, and
we choose~$\nu = 5/2$ and $\sigma^2 = 1$.
We choose three versions of~$\rho_t$: (A) constant across time
at $0.5$, (B) constant within each day but linearly decreasing
across days from~$0.7$ to~$0.3$, and (C) constant within each day but
linearly increasing across days from~$0.3$ to~$0.7$.

to estimate the impact of strong X dependence for close X's then we
will GENERATE SEQUENTIALLY TO FILL IN TO ACCOUNT FOR OVERLAP!
HOW SHOULD WE HANDLE COVARIANCE BETWEEN X's outside the range?
WE WILL APPROXIMATE?

We estimate with~$K_b = K_x = 35$ under both the logistic, penalized
logistic, and multi-level penalized logistic models.

\subsection{Mean-integrated squared error (MISE)}

Since we are primarily interested in accuracy of the final
estimates~$\hat \beta(s)$, we report the mean integrated  squared
error (MISE) of~$\hat \beta (s)$, defined as~$\frac{1}{1000}
\sum_{j=1}^{1000} \int_{-\infty}^t ( \hat \beta_j (t) - \beta(s))^2
ds$ where~$\beta(s) \equiv 0$ for $s > \Delta$.  We define the MISE in
this manner to account for settings where~$\Delta$ is unknown.
For~$\Delta$ known, the MISE simplifies to integrals from~$t-\Delta$
to~$t$.
Table~\ref{} show the MISE of~$\hat \beta (s)$ decomposed into the
variance and squared biase

@@

\begin{proof}[Proof of Lemma~\ref{lemma:simpleasym}]
Lemma~\ref{lemma:simpleasym} will be proven under regularity conditions A-E in~\cite[pp. 420--421]{Andersen1993} along with the following assumptions related to the subsampling rate~$\pi_i (t)$.
We define the joint counting process
\[
M_i (t) = N_i (t) - \int_0^t (h_i (s; \theta) + \pi_i (s)) R_i (s) ds.
\]
where~$N_i (t)$ is the counting process with jumps at all~$t \in
\bfT_i \cup D_i$.

>>>>

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. We use the sandwich smoother~\citep{Xiao2013} to estimate the mean~$\mu_y(s,t) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009} bivariate tensor product splines~\citep{Wood2006} and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(s,t) = X(s,t) - \hat \mu_y(s,t)$ to be the mean-zero process at each time~$t \in \bfT_i \cup \bfD_i$.

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma_y (s, s^\prime) = \int_{0}^\tau c_y( (s,T), (s^\prime, T) ) f_y(T) dt.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c_y((s,t), (s^\prime,t))$ is the covariance function of the windowed covariate history~$X(t,\cdot)$ and $f_y(T)$ is the intensity function for event ($y=1$) and subsampled ($y=0$) times respectively. Estimation of~$\Sigma_y$ occurs in two steps. For simplicity, we present the steps for subsampled times (i.e., $y=0$) but the steps are the same for event times as well. First, the pooled sample covariance is calculated at a set of grid points:
\[
\tilde \Sigma_0 (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | \bfD_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in \bfD_i} \tilde X (s_r,t) \tilde X (s_{r^\prime}, t) \right).
\]
Due to our concern over independent evolution as discussed in section~\ref{section:memproblems}, we do not assume that each observation is observed with white noise and therefore the diagonal
elements of~$\hat \Sigma_0$ are not inflated. Second, the estimator~$\hat \Sigma$ is further smoothed again using the sandwich smoother~\citep{Xiao2013}. Note~\cite{Park2018} smooth the off-diagonal elements, while here we smooth the entire pooled sample covariance matrix. All negative eigenvalues are set to zero to ensure positive semi-definiteness. The result is used as an estimator~$\hat \Sigma_0$ for the pooled covariance $\Sigma_0$.

@@

This connection established by Lemma~\ref{lemma:logistic} between our proposed methodology and logistic regression allows us to leverage ``off-the-shelf'' software.  The main complication is pre-processing of the functional data; however, these additional steps can also be taken care of via existing software.  Therefore, the entire data analytic pipeline is easy-to-implement and requires minimal additional effort by the end-user. To see this, we briefly review the proposed inference procedure.

\begin{rmk}[Inference procedure review] \normalfont
Given observed recurrent event and high frequency data~$\{ \bfT_i, \bfx_i \}_{i=1}^n$,
\begin{enumerate}
\item For each $i \in [n]$, sample non-event times as a time-inhomogeneous Poisson point process with intensity according to~$\pi_i (t)$
\item Estimate mean~$\mu_y (t,s)$ for $0 \leq s \leq \Delta$ at all event times~$t \in \cup_{i=1}^n \bfT_i$ and sampled non-event times~$t \in \cup_{i=1}^n \bfD_i$.
\item Compute marginal covariance across event times,~$\Sigma_1$, and non-event times and~$\Sigma_0$.
\item Compute eigendecomposition~$\{ \hat \psi_k^{(y)}, \hat \lambda_k^{()} \}$ of marginal covariance~$\Sigma_y$
\item Use the eigendecomposition to construct~$W_{i,t}$ for all $i \in [n]$ and $t \in \bfD_i \cup \bfT_i$
\item \label{point:log} Perform logistic regression with binary outcome~$\{ \Y_i (t) \}$  and offset of~$\log \pi_i (t)$.
\end{enumerate}
\end{rmk}

\noindent Before demonstrating the methodology via simulation in Section~\ref{section:simstudy} and a worked example in Section~\ref{section:example}, we provide a theoretical analysis of our current proposal.

@@

Our theoretical analysis requires assumptions regarding the subsampling procedure, the event process, and the functional data. We state these assumptions and then our main theorems. We start by assuming there exists a~$\tau < \infty$ such that all individuals are no longer at risk (i.e.,~$\tau_i < \tau$ for all~$i$). Moreover, define $R_i (t)$ to be the at-risk indicator for participant~$i$, i.e., $R_i (t) = {\bf 1} [t \in (0,\tau_i)]$. Asymptotic theory provided in Lemma~\ref{lemma:simpleasym} will be proven under regularity conditions A-E in~\cite[pp. 420--421]{Andersen1993} along with the following additional assumptions:

\begin{assumption}[Event process assumptions]
\label{assumption:events}\normalfont
We assume the following holds:
\begin{enumerate}[label=(E.\arabic*)]
\item\label{E1} The subsampling rate is both lower and upper bounded for all at-risk times; that is,~$0 < L < \pi_i (t) < U < \infty$ for all~$i=1,2,\ldots$ and~$t \in [0, \tau]$ such that $R_i (t) = 1$.
\item\label{E2} There exists a nonnegative definite matrix~$\Xi (\theta)$ such that
  \[
    n^{-1} \Xi_n (\theta) = n^{-1} \sum_{i=1}^n \int_0^\tau w_i (t;
    \theta) (1 - w_i (t; \theta)) \left[ \frac{h_i^{(1)}(t;
        \theta)}{h_i (t; \theta)} \right] \times  \left [
      \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top R_i (t)
    dt \overset{P}{\to} \Xi (\theta).
  \]
\item\label{E3} There exists~$M$ such that $|W_{i,j,t}| < M$ for
  all~$(i,j,t)$.
\item\label{E4} For all~$j,k$
\[
n^{-1} \sum_{i=1}^n \int_0^\tau \left | \frac{d^2}{d\theta_j
    d\theta_k} h_i (t;\theta_0) \right|^2 R_i (t) dt \overset{P}{\to}
C < \infty
\]
as $n \to \infty$.
\end{enumerate}
\end{assumption}

We also require several assumptions due to the truncation of the Karhunen-Lo{\`e}ve decomposition that represents~$X(s,t)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(s, t) : (s,t) \in \mathcal{S} \times
  \mathcal{T} \}$ is a square integrable element of the $L^2 (
  \mathcal{S} \times \mathcal{T})$
\item\label{A2} The subsampling and conditional intensity rate functions~$f_y(T)$ are continuous and~$\sup |f_y(T)| < \infty$.
\item\label{A3} $\E[X(s,t) X(s^\prime, t) X(s,t^\prime)
  X(s^\prime, t^\prime) ] < \infty$ for
  each $s,s^\prime \in [0,\Delta]$ and~$0 < t, t^\prime < \tau$.
\item\label{A4} $\E[\|X(\cdot, t)\|^4] < \infty$ for each~$0< t < \tau$.
\end{enumerate}
\end{assumption}
Finally, for simplicity, we assume that there exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the true function~$\beta(t)$ sits in the span of the spline basis expansion. With the necessary assumptions stated, we can now state Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.
\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events},
Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for
large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
An estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i}
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process, standard logistic regression software can be used to fit the recurrent event model by specifying an offset equal to~$\log \pi_i (t)$. Not only this, Lemma~\ref{prop:optimal} shows weights~\eqref{eq:waage_weights} are optimal within a particular class of weighted estimating equations.

\begin{lemma} \normalfont
\label{prop:optimal}
If the event process is an inhomogeneous Poisson point process with intensity~$h(t; \theta)$ and subsampling occurs via an independent, inhomogeneous Poisson point process with intensity~$\pi (t)$, then~$\hat U_n (\theta)$ are optimal estimating functions (i.e., most efficient) in the class of weighted estimating functions given by~\eqref{eq:approxscore} replacing~\eqref{eq:waage_weights} by any weight function~$w_i (t; \theta)$. This class includes the Horvitz-Thompson estimator under~$w(s; \theta) = 1$.
\end{lemma}

\noindent Proposition~\ref{prop:optimal} ensures the only loss of statistical efficiency is due to subsampling and not using a suboptimal estimation procedure given subsampling.
% This result follows from the probability-based sampling design literature~\cite{Rathbun2012}.

@@

Under assumption~\ref{assumption:truncation}, we next consider the statistical efficiency of our proposed estimator when compared to complete-data maximum likelihood estimation. While subsampling introduces additional variation, it may significantly reduce the overall computational burden. It is this trade-off that we next make precise. To start, we consider the subsampling rate~$\pi(t) = c \times h(t; \theta)$ for $c>0$, i.e., the subsampling rate proportional to the intensity function with time-independent constant~$c > 0$. Under this subsampling rate, the weight function~\eqref{eq:waage_weights} is equal to $c / (c+1)$. Under Lemma~\ref{lemma:simpleasym},
\[
\Xi (\theta) = \frac{c}{c+1} \int_0^\tau \frac{ h^{(1)} (t; \theta)
  h^{(1)} (t; \theta)^\top}{h (t; \theta)} dt = \frac{c}{c+1} \Sigma (\theta)
\]
where~$\Sigma(\theta)$ is the Fisher information of the complete-data maximum likelihood estimator.
Therefore the relative efficiency is~$c/(1+c)$. For an upper bound~$H = \max_{t \in (0,\tau)} h(t;\theta)$, if we set~$\pi (t) = c \times H$, then the relative efficiency can be lower bounded by~$c / (c+1)$.

Recall sensor measurements occur on the order of multiple times per second.  Suppose the intensity rate is bounded above by $3$ and the unit time scale was hours. If we can subsample the data at~$30$ times per hour, then we have a lower bound on the efficiency of $0.909$. For a 4Hz sensor, this reduces the number of samples per hour from~$4 \times 60 \times 60 = 14,400$ per hour to on average $30$. While the computational complexity of logistic regression is linear in the number of samples, we get $480$ times reduction in the data size at the cost of a $0.909$ statistical efficiency. If we sample~$300$ times per hour, then the efficiency loss is only $0.999$, with a $48$ times reduction in data size. Table~\ref{tab:compvseff} provides additional examples for a 4Hz sensor.  The data reduction depends on the sensor measurement rate; however, the lower bound on statistical efficiency does not.  This is because the subsampling rate only depends on the upper bound of the intensity function. That is, if the events are rare then the amount of data we need to subsample is greatly reduced with no impact to statistical efficiency.

@@

Recall theoretical results were proven under the assumption that there exists~$b^\star$ such that~$\beta(t) = \phi(t) {\bf b}^\star$. To make this assumption plausible, we set~$K_b$ large enough (but less than~$K_x$ to ensure identifiability) to ensure the spline basis expansion is sufficiently expressive. However, in practice, such a choice of~$K_b$ may lead to overfitting the data. Following~\cite{Goldsmith2011}, we choose the polynomial spline model and set~$\phi(t) = b_0 + b_1 t + b_2 t^2 + \sum_{j=3}^{K_b} (t-\kappa_j)^3$ where~$\{ \kappa_j \}_{j=3}^{K_b}$ are the chosen knots and assume~$\{ b_j \}_{j=3}^{K_b} \sim N(0, \sigma^2 I)$ to induce smoothness on the spline model.  Combining the penalized spline formulation with Lemma~\ref{lemma:logistic} establishes a connection between our approximate score equations and solving a generalized mixed effects logistic regression with offset.

@@

Due to the connection with generalized mixed effects models, we can leverage existing inferential machinery to obtain variance-covariance estimates of model parameters. That is, if~$\hat \Sigma_{bb}$ is the $K_b \times K_b$ dimensional matrix obtained by plugging in the estimates of variance components into the formula for the variance of~$\hat b$, then the standard error for estimate at time~$t_0$ -- i.e., ~$\hat \beta (t_0) = \phi(t_0) \hat {\bf b}$ -- is given by~$\sqrt{ \phi (t_0 ) \hat \Sigma_{bb} \phi(t_0)^\top}$.  Then the approximate 95\% confidence interval can be constructed as~$\hat \beta (t_0) \pm 1.96 \sqrt{\phi (t_0) \hat \Sigma_{bb} \phi(t_0)^\top}$.

@@

We next assess the proposed methodology via a simulation study. Here, we assume each individual is observed over five days each defined on the unit interval~$[0,1]$ with $1000$ equally spaced observation times per day. We define~$X(t)$ at the grid of observerations as a mean-zero Gaussian process with covariance
\[
  \Sigma (t, t^\prime) =
  \frac{\sigma^2}{\Gamma (v) 2^{\nu-1}}
  \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)^{\nu}
  K_v \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)
\]
where~$K_v$ is the modified Bessel function of the second kind.  We set~$\nu = 1/2$, $\sigma^2 = 1$, and~$\rho = 0.5$ as well as set~$K_b = K_x = 35$.  For simplicity, we assume~$\Sigma$ is known in computation of the the eigendecomposisions. Given~$\{ X(t) \}_{0 \leq t \leq 1}$ for a given user-day, we generate event times according a chosen hazard function~$h (t; theta)$.  To mimic our real data, we set
\[
h(t; \theta) = \exp \left( \theta_0 + \int_{0}^{\Delta} X(t-s) \beta(s) ds \right).
\]
We set~$\Delta = 1/(12*2)$ to mimic a $30$-minute window for a $12$-hour day.  We set~$\theta_0 = \log(5/1000)$ to set a baseline risk of approximately 5 events per day. We consider two choices of~$\beta(s)$: (1) $\beta_0 + \exp(- \beta_1 s) $, and (2) $\beta_1  * \sin \left( 2 \pi \frac{s}{\Delta} - \pi/2 \right)$

We sample non-events at rates of 1 per 12/6/1/0.5 hours.

Since we are primarily interested in accuracy of the final estimates~$\hat \beta(s)$, we report the mean integrated squared error (MISE) of~$\hat \beta (s)$, defined as~$\frac{1}{100} \sum_{j=1}^{100} \int_{0}^{\Delta^\prime} ( \hat \beta_j (s) - \beta(s))^2 ds$ where~$\beta(s) \equiv 0$ for $s > \Delta$.  We define the MISE in this manner to account for settings where~$\Delta$ is unknown. Table~\ref{tab:mise} show the MISE decomposed into the variance and squared biase

\begin{table}[!th]
\begin{tabular}{c | c c c c c}
Sampling rate & 0.5 & 1 & 5 & 25 & 50 \\
Variance
Squared Bias
MISE
\end{tabular}
\caption{XXX}
\label{tab:mise}
\end{table}

@@

\begin{proof}[Proof of Lemma~\ref{lemma:simpleasym}]
% Lemma~\ref{lemma:simpleasym} will be proven under regularity conditions A-E in~\cite[pp. 420--421]{Andersen1993} along with the following assumptions related to the subsampling rate~$\pi_i (t)$.
We define the joint counting process
\[
M_i (t) = N_i (t) - \int_0^t (h_i (s; \theta) + \pi_i (s)) R_i (s) ds.
\]
where~$N_i (t)$ is the counting process with jumps at all~$t \in \bfT_i \cup \bfD_i$. Asymptotic consistency is guaranteed~\cite[Theorem VI.1.1]{Andersen1993} if
\begin{equation}
\label{eq:andersen1}
n^{-1} \tilde U_n (\theta_0 ) \overset{P}{\to} 0,
\end{equation}
and
\begin{equation}
\label{eq:andersen2}
n^{-1} \frac{d}{d \theta^{\top}} \tilde U_n (\theta) \big |_{\theta =
  \theta_0} = \Xi (\theta),
\end{equation}
and
\begin{equation}
\label{eq:andersen3}
\lim_{n \to \infty} P \left( \left | n^{-1} \frac{d^2}{d \theta_j d
      \theta_k} \tilde U_n (\theta) \right | < M \text{ for all }
  j,k \text{ and all } \theta \in \Theta_0 \right) = 1
\end{equation}
To prove~\eqref{eq:andersen1}, we decompose~$\tilde U_n (\theta_0)$ into
three terms
\[
n^{-1} \tilde U_n (\theta_0) = n^{-1} U_n (\theta_0 )  + n^{-1} \left(
  \hat U_n (\theta_0) - U_n (\theta) \right) + n^{-1} \left(\tilde U_n
  (\theta_0) - \hat U_n (\theta_0) \right)
\]
where, recall,~$\hat U_n (\theta_0)$ are the logistic score equations with $\int_0^\infty X_i(t-s) \beta(s) ds$ and~$\tilde U_n (\theta_0)$ with the approximation~$C_{i,t}^\top J_{\phi, \psi} {\bf b}$. \cite{Andersen1993} (Theorem VI.1.1) show that~$n^{-1} U_n (\theta_0)
\to 0$. The second term
\begin{align*}
n^{-1} \left( \hat U_n (\theta_0) - U_n (\theta_0) \right) =
  \frac{1}{n} \sum_{i=1}^n \int_0^{\tau} \frac{h^{(1)} (s;
  \theta_0)}{\pi_i (s) + h (s;\theta_0)} d M_i (s).
\end{align*}
where~$h(\cdot; \theta_0)$ is given by~\eqref{eq:hazardlinear}. Lenglart's inequality implies the second term converges in probability to zero under Assumptions~\ref{E1} and~\ref{E2}. The third term satisfies
\begin{align*}
  &n^{-1} \left \| \hat U_n (\theta_0) - \tilde U_n (\theta_0) \right \| \\
